Simple repression draft

### 2020-05-20 doc/section_02_means.tex
RP report on May 19 morning
wrote abstract
went carefully through section 1 on introduction
went carefully through section 2 on means
addded many items to bib file

% !TEX root = ./busty_transcription.tex
\section{Means}
\subsection{Fold-changes are indistinguishable across models}
As a first stop on our search for the ``right'' model of simple repression, let
us consider what we can learn from theory and measurements of fold-change in
mean expression. The fold-change is defined as
\begin{equation}
\text{fold-change}
= \frac{\langle \text{gene expression with repressor present} \rangle}
        {\langle \text{gene expression with repressor absent} \rangle}
= \frac{\langle m\rangle(R\ne 0)}{\langle m\rangle(R = 0)}
\label{eq:fc_def}
\end{equation}
where angle brackets denote the averge over a population of cells and mean mRNA
$\langle m\rangle$ is viewed as a function of repressor copy number $R$. The
second equality in~\ref{eq:fc_def} follows from assuming that the translation
efficiency, i.e., the number of proteins translated per mRNA, is the same in
both conditions. In other words, we assume that mean protein level is
proportional to mean mRNA level, and that the proportionality constant is the
same in both conditions and therefore cancels out in the ratio. This is
reasonable since the cells in the two conditions are identical except for the
presence of the transcription factor, and it is difficult to imagine the
transcription factor having any direct effect on translation.

Fold-change has proven a very convenient observable in past
work~\cite{Garcia2011a, Razo-Mejia2018, Chure2019}. Part of its utility in
dissecting transcriptional regulation is its ratiometric nature, which is always
easier than making an absolute measurement. Also, by measuring otherwise
identical cells with and without a transcription factor present, any biological
noise common to both conditions can be made to cancel away. In addition, the
presence of different operator sites is known to change the basal expression
level in the absence of repressor~\mmnote{look up cite}. Computing fold-changes
neatly cancels this effect as well.

\fig{fig1:means_cartoons} depicts a smorgasbord of mathematized cartoons for
simple repression that have appeared in previous literature. For each cartoon,
we calculate the fold-change in mean gene expression as predicted by that model,
deferring some algebraic details to the appendix. What we will find is that all
cartoons collapse to a single master curve, shown in \fig{fig1:means_cartoons}c,
which contains just two parameters. We label the parameters $\Delta F_R$, an
effective free energy parametrizing the repressor-DNA interaction, and $\rho$,
which subsumes all details of transcription in the absence of repressors. We
will offer some intuition for why this master curve exists and discuss why it is
likely impossible to discriminate ``right'' from ``wrong'' cartoons given only
measurements of fold-changes in expression.

\subsubsection{Equilibrium, 2-state}
In this simplest model, the promoter is idealized as existing in one of two
states, either repressor bound or repressor unbound. The rate of transcription,
and therefore the level of gene expression, is assumed to be proportional to the
fraction of time spent in the repressor unbound state. Mathematically, this
amounts to the statement that the mean mRNA level is given by
\begin{equation}
\langle m \rangle = \frac{r}{\gamma}
        \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}\right)^{-1},
\end{equation}
where $r$ is the transcription rate from the repressor unbound state, $\gamma$
is the mRNA degradation rate, $R$ is repressor copy number, $N_{NS}$ is the
number of nonspecific binding sites in the genome (where repressors spend most
of their time when not bound to the operator, their specifc binding site),
$\beta=1/k_BT$, and $\Delta\epsilon_R$ is the binding energy of a repressor to
its operator site. The derivation of this result is deferred to the appendix.

The fold-change is easily found as the ratio of mean mRNA with and without
repressor. This gives
\begin{equation}
\text{fold-change}
= \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}\right)^{-1},
\end{equation}
which clearly matches the form of the master curve in~\eq{fig1:means_cartoons}
with $\rho=1$.

In fact it was noted in~\cite{Chure2019} that this two-state model can be viewed
as the coarse-graining of any equilibrium promoter model in which no
transcriptionally active states have transcription factor bound. Put
differently, there is no overlap between TF bound states and transcriptionally
active states. We will see this explicitly in the 3-state equilibrium model
below, but perhaps surprising is that something analogous carries over even to
the nonequilibrium models we consider later.

\subsubsection{Equilibrium, 3-state}
Compared to the previous model, here we fine-grain the repressor unbound state
into two separate states: empty, and RNAP bound. This picture was used
in~\cite{Garcia2011a} as we use it here, and in~\cite{Razo-Mejia2018}
and~\cite{Chure2019} it was generalized to incorporate small-molecule inducers
that bind the repressor. Interestingly, the effect of this is simply to
renormalize $R$ from the total number of repressors to a smaller effective
number, roughly (though not exactly) the number which are unbound by inducers.

The mean mRNA, as derived in the appendix from a similar enumeration of states
and weights as the previous model, is
\begin{equation}
\langle m \rangle = \frac{r}{\gamma}
\frac{\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
        {
        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}
        + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}
        },
\end{equation}
where the new variables are $\Delta\epsilon_P$, the difference in RNAP binding
energy to its specific site (the promoter) relative to an average nonspecific
background site, and the RNAP copy number, $P$. The fold-change again follows
immediately as
\begin{align}
\text{fold-change}
&= \frac{\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
        {
        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}
        + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}
        }
\frac{1 + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
        {\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
\\
&= \left(
1 + \frac{\frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}}
        {1 + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
\right)^{-1}
\\
&= (1 + \exp(-\Delta F_R - \log\rho))^{-1},
\end{align}
with $\Delta F_R = \beta\Delta\epsilon_R - \log(R/N_{NS})$ and $\rho = 1 +
\frac{P}{N_{NS}}\mathrm{e}^{-\beta\Delta\varepsilon_P}$ as shown
in~\fig{fig1:means_cartoons}.

\begin{figure}%[h!]
\centering
\includegraphics[width=\textwidth]{../figures/fig1/fig1.ai}
\caption{\textbf{An overview of the simple repression motif at the level of
means.} (A) depicts the qualitative biological picture of simple repression. (B)
shows a variety of possible mathematized cartoons of simple repression, along
with the effective parameter $\rho$ which subsumes all regulatory details of the
architecture excluding the repressor. (C) shows the ``master curve'' to which
all cartoons in (B) collapse.}
\label{fig1:means_cartoons}
\end{figure}

\subsubsection{Nonequilibrium model 1 - Poisson promoter}
For our first kinetic model, we consider the simplest possible picture with only
two states, repressor bound and unbound. This is exactly the model used for the
main results of~\cite{Jones2014}. In this picture, repressor association and
dissociation rates from its operator site, $k_R^+$ and $k_R^-$, respectively,
govern transitions between the two states. When the system is in the unbound
state, transcriptions initiate at rate $r$, which represents a coarse-graining
of all the downstream processes into a single effective rate. mRNA is again
degraded at rate $\gamma$.

Let $p_{m,R}(t)$ denote the joint probability of finding the system in the
repressor bound state with $m$ mRNA molecules present at time $t$, and similarly
define $p_{m,U}(t)$ for the repressor unbound state. This model is governed by a
master equation giving the time evolution of $p_{m,R}(t)$ and $p_{m,U}(t)$,
which is
\begin{align}
\begin{split}
\deriv{t}p_{m,R}(t) =& - k_R^- p_{m,R}(t) + k_R^+ p_{m,U}(t)
                + (m+1)\gamma p_{m+1,R}(t) - \gamma p_{m,R}(t)
\\
\deriv{t}p_{m,U}(t) =&\; k_R^- p_{m,R}(t) - k_R^+ p_{m,U}(t)
                        + rp_{m-1,U}(t) - rp_{m,U}(t)
                \\
                &+ (m+1)\gamma p_{m+1,U}(t) - \gamma p_{m,U}(t).
\label{eq:poisson_promoter_cme}
\end{split}
\end{align}
In each equation, the first two terms describe transitions between promoter
states due to repressors unbinding and binding, respectively. The final two
terms describe degradation of mRNA, decreasing the copy number by one, and the
terms with coefficient $r$ describe transcription initiation increasing the mRNA
copy number by one.

We can greatly simplify the notation, which will be especially useful for the
more complicated models yet to come, by re-expressing the master equation in
vector form. The promoter states are collected into a vector and the rate
constants are collected into matrices as
\begin{equation}
\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,U} \end{pmatrix},\
\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ \\ k_R^- & -k_R^+ \end{pmatrix},\
\matr{R} = \begin{pmatrix} 0 & 0 \\ 0 & r \end{pmatrix},\
\label{eq:2state_cme_matrices}
\end{equation}
so that the master equation may be condensed as
\begin{equation}
\deriv{t}\vect{p}_m(t) =
\left( \matr{K} - \matr{R} - \gamma m \matr{I} \right) \vect{p}_m(t)
                + \matr{R} \vect{p}_{m-1}(t)
                + \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
\label{eq:2state_rep_cme}
\end{equation}
where $\matr{I}$ is the identity matrix. Taking steady state by setting time
derivatives to zero, the mean mRNA can be found to be
\begin{equation}
\langle m \rangle = \frac{r}{\gamma}
        \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
\end{equation}
with the algebra details again deferred to the appendix. Recall $k_R^+$ is
proportional to the repressor copy number, so in computing fold-change, absence
of repressor corresponds to $k_R^+\rightarrow0$. Therefore fold-change in this
model is simply
\begin{equation}
\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
\end{equation}
again matching the master curve of~\fig{fig1:means_cartoons} with $\rho=1$.

\subsubsection{Nonequilibrium model 2 - RNAP bound/unbound states}
Analogous to the three-state equilibrium model, this model fine-grains the
repressor unbound state of nonequilibrium model 1, resolving it into an empty
promoter state and an RNAP-bound state. Note in this picture, in contrast with
model 4 below, transcription initiation is accompanied by a promoter state
change, in keeping with the interpretation as RNAP-bound and empty states: if an
RNAP successfully escapes the promoter and proceeds to elongation of a
transcript, clearly it is no longer bound at the promoter! Therefore another
RNAP must bind before another transcript can be initiated.

The master equation governing this model is analogous
to~\eqrange{eq:2state_cme_matrices}{eq:2state_rep_cme} for model 1 above. The
main subtlety arises since transcription initiation accompanies a promoter state
change. This can be understood by analogy to $\matr{K}$. The off-diagonals and
diagonals of $\matr{K}$ correspond to transitions arriving at or departing from,
respectively, the promoter state of interest. If transcription initiations are
accompanied by promoter state changes, we must have separate matrices for
arriving and departing transcription events since the arriving and departing
transitions have different initial copy number of mRNA, unlike for $\matr{K}$
where they are the same. The master equation for this model is
\begin{equation}
\deriv{t}\vect{p}_m(t) =
\left( \matr{K} - \matr{R_D} - \gamma m \matr{I} \right) \vect{p}_m(t)
                + \matr{R_A} \vect{p}_{m-1}(t) +
                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
\label{eq:3state_rep_cme}
\end{equation}
with the state vector and promoter transition matrix defined as
\begin{equation}
\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\ p_{m,P} \end{pmatrix},\
\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 \\
                        k_R^- & -k_R^+ -k_P^+ & k_P^- \\
                        0 & k_P^+ & -k_P^- 
                \end{pmatrix},
\label{eq:3state_cme_matrices_pt1}
\end{equation}
and the initiation matrices given by
\begin{equation}
\matr{R_A} = \begin{pmatrix}
                0 & 0 & 0 \\ 0 & 0 & r \\ 0 & 0 & 0
                \end{pmatrix},\
\matr{R_D} = \begin{pmatrix}
                0 & 0 & 0 \\ 0 & 0 & 0 \\ 0 & 0 & r
                \end{pmatrix},
\label{eq:3state_cme_matrices_pt2}
\end{equation}
The elements of $\vect{p}_m$ encode the probabilities of having $m$ mRNA present
along with the promoter having repressor bound, being empty, or having RNAP
bound, respectively. $\matr{R_A}$ describes probability flux \textit{arriving}
at the state $\vect{p}_m$ from a state with one fewer mRNA, namely
$\vect{p}_{m-1}$, and $\matr{R_D}$ describes probability flux \textit{departing}
from the state $\vect{p}_m$ for a state with one more mRNA, namely
$\vect{p}_{m+1}$. $\matr{K}$ is closely analogous to model 1.

Mean mRNA at steady state is found analogously to model 1, with the result
\begin{equation}
\langle m\rangle = \frac{r}{\gamma}
        \frac{k_R^- k_P^+}
        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)}.
\end{equation}
Fold-change is again found from the ratio prescribed by~\eq{eq:fc_def}, from
which we have
\begin{align}
\text{fold-change}
&=      \frac{k_R^- k_P^+}
        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)}
        \frac{k_P^+ + k_P^- + r}{k_P^+}
\\
&=      \left(1 + \frac{k_R^+}{k_R^-}
                \frac{k_P^- + r}{k_P^+ + k_P^- + r}
        \right)^{-1}
\\
&=      \left(1 + \frac{k_R^+}{k_R^-}
        \left(1 + \frac{k_P^+}{k_P^- + r}\right)^{-1}
        \right)^{-1},
\end{align}
which follows the master curve with $\rho = 1 + k_P^+/(k_P^- + r)$ as claimed.

\subsubsection{Nonequilibrium model 3 - Multistep transcription initiation and
escape} One might reasonably complain that the first two ``nonequilibrium''
models we have considered are straw men. Their steady states necessarily satisfy
detailed balance which is equivalent to thermodynamic equilibrium. Why is this
the case? At steady state there is by definition no net probability flux in or
out of each promoter state, but since the promoter states form a linear chain,
there is only one way in or out of the repressor bound and RNAP bound states,
implying each edge must actually have a net zero probability flux, which is the
definition of detailed balance (usually phrased as equality of forward and
reverse transition fluxes).

Now we consider model 3 in~\eq{fig1:means_cartoons} which allows the possibility
of true nonequilibrium steady-state fluxes through the promoter states. We point
out that this model was considered previously in~\cite{Mitarai2015} where a
comparison was made with model 1 as used in~\cite{Jones2014}. The authors
of~\cite{Mitarai2015} argued that the additional complexity is essential to
properly account for the noise in the mRNA distribution. We will weigh in on
both models later when we consider observables beyond fold-change.

The master equation governing this model is identical in form to model 2 above,
namely
\begin{equation}
\deriv{t}\vect{p}_m(t) =
\left( \matr{K} - \matr{R_D} - \gamma m \matr{I} \right) \vect{p}_m(t)
                + \matr{R_A} \vect{p}_{m-1}(t) +
                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
\end{equation}
but obviously with a higher-dimensional state space and different matrices. The
state vector and promoter transition matrix are now
\begin{equation}
\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\
                             p_{m,C} \\ p_{m,O}\end{pmatrix},\
\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 & 0\\
                        k_R^- & -k_R^+ -k_P^+ & k_P^- & 0 \\
                        0 & k_P^+ & -k_P^- - k_O & 0 \\
                        0 & 0 & k_O & 0
                \end{pmatrix},
\end{equation}
with the four promoter states, in order, being repressor bound, empty, RNAP
closed complex, and RNAP open complex. Besides increasing dimension by one
again, the only new feature in $\matr{K}$ is the rate $k_O$, representing the
(irreversible) rate of open complex formation from the closed complex. The
initiation matrices are given by
\begin{equation}
\matr{R_A} = \begin{pmatrix}
        0 & 0 & 0 & 0 \\ 0 & 0 & 0 & r \\ 0 & 0 & 0 & 0\\ 0 & 0 & 0 & 0
                \end{pmatrix},\
\matr{R_D} = \begin{pmatrix}
        0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0\\ 0 & 0 & 0 & r
                \end{pmatrix},
\end{equation}
again closely analogous to model 2.

>>>>

% !TEX root = ./busty_transcription.tex
\section{Means}

As noted in the previous section, there are two broad classes of models
in play for computing the input-output functions of regulatory architectures 
as shown in Figure~\ref{fig1:means_cartoons}.
In both classes of model, the promoter is imagined to exist in a discrete
set of states of occupancy, with each such state of occupancy
accorded its own rate of transcription (and for many of these
states, there is no transcription).  The models are probabilistic with
each state assigned some probability and the overall rate of transcription
given by 
\begin{equation}
\mbox{average rate of transcription} = \sum_i r_i p_i,
\end{equation}
where $i$ labels the distinct states, $p_i$ is the probability of
the i$^{th}$ state and $r_i$ is the rate of transcription of
that state.  Ultimately, the different models differ along several key
axes: what states to consider and how to compute the probabilities of those
states.

The first class of models that are the focus of
the present section on predicting the mean
level of gene expression, sometimes known as thermodynamic models,
invoke the tools of equilibrium statistical mechanics to compute the 
probabilities~\cite{Ackers1982, Shea1985, Buchler2003,Vilar2003a,Vilar2003b, Bintu2005a,Bintu2005c, Gertz2009,Sherman2012, Saiz2013}.
As seen in Figure~\ref{fig1:means_cartoons}(B), even within
the class of thermodynamic models, we can make different commitments
about the underlying microscopic states of the promoter.  Indeed,
the list of options considered here does not at all exhaust the suite
of different microscopic states we can assign to the promoter.

The second class of models that allow us to access the mean
expression use chemical master equations to compute the
probabilities of the different microscopic states
~\cite{Ko1991,Peccoud1995,Record1996,
Kepler2001, Sanchez2008,Shahrezaei2008,Sanchez2011, Michel2010}.
As seen in Figure~\ref{fig1:means_cartoons}(B), we consider a host of
different nonequilibrium models, each of which will have its own 
result for both the mean and noise in gene expression.

%The second class of models based on chemical master equations compute the time-evolution
%of the probability distribution for mRNA.

\subsection{Fold-changes are indistinguishable across models}
As a first stop on our search for the ``right'' model of simple repression, let
us consider what we can learn from theory and measurements on the
average level of  gene expression in a population of cells.
One experimental strategy that has been particularly useful (if incomplete since
it misses out on gene expression dynamics) is to measure the  fold-change in
mean expression. The fold-change is defined as
\begin{equation}
\text{fold-change}
= \frac{\langle \text{gene expression with repressor present} \rangle}
        {\langle \text{gene expression with repressor absent} \rangle}
= \frac{\langle m\rangle(R\ne 0)}{\langle m\rangle(R = 0)}
\label{eq:fc_def}
\end{equation}
where angle brackets denote the average over a population of cells and mean mRNA
$\langle m\rangle$ is viewed as a function of repressor copy number $R$. The
second equality in~\ref{eq:fc_def} follows from assuming that the translation
efficiency, i.e., the number of proteins translated per mRNA, is the same in
both conditions. In other words, we assume that mean protein level is
proportional to mean mRNA level, and that the proportionality constant is the
same in both conditions and therefore cancels out in the ratio. This is
reasonable since the cells in the two conditions are identical except for the
presence of the transcription factor, and the model assumes that the 
transcription factor has no direct effect on translation.

Fold-change has proven a very convenient observable in past
work~\cite{Garcia2011a, Razo-Mejia2018, Chure2019}. Part of its utility in
dissecting transcriptional regulation is its ratiometric nature, which removes many
secondary effects that are present when making an absolute gene
expression measurement.   Also, by measuring otherwise
identical cells with and without a transcription factor present, any biological
noise common to both conditions can be made to cancel away. In addition, the
presence of different operator sites is known to change the basal expression
level in the absence of repressor~\mmnote{look up cite}. Computing fold-changes
neatly cancels this effect as well.

\fig{fig1:means_cartoons} depicts a smorgasbord of mathematicized cartoons for
simple repression using both thermodynamic and kinetic models  that have appeared in previous literature. For each cartoon,
we calculate the fold-change in mean gene expression as predicted by that model,
deferring some algebraic details to the appendix. What we will find is that all
cartoons collapse to a single master curve, shown in \fig{fig1:means_cartoons}(C),
which contains just two parameters. We label the parameters $\Delta F_R$, an
effective free energy parametrizing the repressor-DNA interaction, and $\rho$,
which subsumes all details of transcription in the absence of repressors. We
will offer some intuition for why this master curve exists and discuss why at the level of
the mean expression, we are unable   to discriminate ``right'' from ``wrong'' cartoons given only
measurements of fold-changes in expression.

\subsubsection{The Two-State Equilibrium Model}
In this simplest model, the promoter is idealized as existing in one of two
states, either repressor bound or repressor unbound. The rate of transcription,
and therefore the level of gene expression, is assumed to be proportional to the
fraction of time spent in the repressor unbound state. Mathematically, this
amounts to the statement that the mean mRNA level is given by
\begin{equation}
\langle m \rangle = \frac{r}{\gamma}
        \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}\right)^{-1},
\end{equation}
where $r$ is the transcription rate from the repressor unbound state, $\gamma$
is the mRNA degradation rate, $R$ is repressor copy number, $N_{NS}$ is the
number of nonspecific binding sites in the genome (where repressors spend most
of their time when not bound to the operator, their specifc binding site),
$\beta=1/k_BT$, and $\Delta\epsilon_R$ is the binding energy of a repressor to
its operator site. The derivation of this result is deferred to the appendix.

The fold-change is  found as the ratio of mean mRNA with and without
repressor as introduced in eqn.~\ref{eq:fc_def}. Invoking that definition results in
\begin{equation}
\text{fold-change}
= \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}\right)^{-1},
\end{equation}
which clearly matches the form of the master curve in Figure~\ref{fig1:means_cartoons}(C)
with $\rho=1$.

In fact it was noted in~\cite{Chure2019} that this two-state model can be viewed
as the coarse-graining of any equilibrium promoter model in which no
transcriptionally active states have transcription factor bound. Put
differently, there is no overlap between TF bound states and transcriptionally
active states. We will see this explicitly in the 3-state equilibrium model
below, but perhaps surprising is that an analogous result carries over even to
the nonequilibrium models we consider later.

\subsubsection{The Three-State Equilibrium Model}
Compared to the previous model, here we fine-grain the repressor unbound state
into two separate states: empty, and RNAP bound. This picture was used
in~\cite{Garcia2011a} as we use it here, and in~\cite{Razo-Mejia2018}
and~\cite{Chure2019} it was generalized to incorporate small-molecule inducers
that bind the repressor. Interestingly, the effect of this is simply to
renormalize $R$ from the total number of repressors to a smaller effective
number, roughly (though not exactly) the number which are unbound by inducers.

The mean mRNA, as derived in the appendix from a similar enumeration of states
and weights as the previous model, is
\begin{equation}
\langle m \rangle = \frac{r}{\gamma}
\frac{\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
        {
        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}
        + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}
        },
\end{equation}
where the new variables are $\Delta\epsilon_P$, the difference in RNAP binding
energy to its specific site (the promoter) relative to an average nonspecific
background site, and the RNAP copy number, $P$. The fold-change again follows
immediately as
\begin{align}
\text{fold-change}
&= \frac{\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
        {
        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}
        + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}
        }
\frac{1 + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
        {\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
\\
&= \left(
1 + \frac{\frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}}
        {1 + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
\right)^{-1}
\\
&= (1 + \exp(-\Delta F_R - \log\rho))^{-1},
\end{align}
with $\Delta F_R = \beta\Delta\epsilon_R - \log(R/N_{NS})$ and $\rho = 1 +
\frac{P}{N_{NS}}\mathrm{e}^{-\beta\Delta\varepsilon_P}$ as shown
in~\fig{fig1:means_cartoons}.  Thus far,
we see that the two thermodynamic models, despite making
different coarse-graining commitments, result in the same functional
form for the fold-change in mean gene expression.  We now
explore how kinetic models fare when faced with computing
the same observable.

\subsubsection{The Poisson Promoter Nonequilibrium Model}
For our first kinetic model, we  imitate the states considered in
the Two-State Equilibrium Model and consider the simplest possible picture with only
two states, repressor bound and unbound. This is exactly the model used for the
main results of~\cite{Jones2014}. In this picture, repressor association and
dissociation rates from its operator site, $k_R^+$ and $k_R^-$, respectively,
govern transitions between the two states. When the system is in the unbound
state, transcription initiates at rate $r$, which represents a coarse-graining
of all the downstream processes into a single effective rate. mRNA is 
degraded at rate $\gamma$ as already exploited in the previous models.

Let $p_{m,R}(t)$ denote the joint probability of finding the system in the
repressor bound state with $m$ mRNA molecules present at time $t$, and similarly
define $p_{m,U}(t)$ for the repressor unbound state. This model is governed by coupled
master equations giving the time evolution of $p_{m,R}(t)$ and 
$p_{m,U}(t)$~\cite{Sanchez2008,Sanchez2011, Phillips2019}
which we can write as 
\begin{align}
\begin{split}
\deriv{t}p_{m,R}(t) =& - k_R^- p_{m,R}(t) + k_R^+ p_{m,U}(t)
                + (m+1)\gamma p_{m+1,R}(t) - \gamma p_{m,R}(t)
\\
\deriv{t}p_{m,U}(t) =&\; k_R^- p_{m,R}(t) - k_R^+ p_{m,U}(t)
                        + rp_{m-1,U}(t) - rp_{m,U}(t)
                \\
                &+ (m+1)\gamma p_{m+1,U}(t) - \gamma p_{m,U}(t),
\label{eq:poisson_promoter_cme}
\end{split}
\end{align}
where each term on the right corresponds to a transition between two
states of the promoter.
In each equation, the first two terms describe transitions between promoter
states due to repressors unbinding and binding, respectively. The final two
terms describe degradation of mRNA, decreasing the copy number by one, and the
terms with coefficient $r$ describe transcription initiation increasing the mRNA
copy number by one.

We can greatly simplify the notation, which will be especially useful for the
more complicated models yet to come, by re-expressing the master equation in
vector form~\cite{Phillips2012}. The promoter states are collected into a vector and the rate
constants are collected into matrices as
\begin{equation}
\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,U} \end{pmatrix},\
\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ \\ k_R^- & -k_R^+ \end{pmatrix},\
\matr{R} = \begin{pmatrix} 0 & 0 \\ 0 & r \end{pmatrix},\
\label{eq:2state_cme_matrices}
\end{equation}
so that the master equation may be condensed as
\begin{equation}
\deriv{t}\vect{p}_m(t) =
\left( \matr{K} - \matr{R} - \gamma m \matr{I} \right) \vect{p}_m(t)
                + \matr{R} \vect{p}_{m-1}(t)
                + \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
\label{eq:2state_rep_cme}
\end{equation}
where $\matr{I}$ is the identity matrix. Taking steady state by setting time
derivatives to zero, the mean mRNA can be found to be
\begin{equation}
\langle m \rangle = \frac{r}{\gamma}
        \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
\end{equation}
with the algebra details again deferred to the appendix. Recall $k_R^+$ is
proportional to the repressor copy number, so in computing fold-change, absence
of repressor corresponds to $k_R^+\rightarrow0$. Therefore fold-change in this
model is simply
\begin{equation}
\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
\end{equation}
again matching the master curve of~\fig{fig1:means_cartoons}(C) with $\rho=1$.

\subsubsection{Nonequilibrium Model Two - RNAP Bound and Unbound States}
Our second kinetic model mirrors the second equilibrium model of
Figure~\ref{fig1:means_cartoons}(B) by fine-graining  the
repressor unbound state of nonequilibrium model 1, resolving it into an empty
promoter state and an RNAP-bound state. Note in this picture, in contrast with
model 4 below, transcription initiation is accompanied by a promoter state
change, in keeping with the interpretation as RNAP-bound and empty states: if an
RNAP successfully escapes the promoter and proceeds to elongation of a
transcript, clearly it is no longer bound at the promoter! Therefore another
RNAP must bind before another transcript can be initiated.

The master equation governing this model is analogous
to~\eqrange{eq:2state_cme_matrices}{eq:2state_rep_cme} for model 1 above. The
main subtlety arises since transcription initiation accompanies a promoter state
change. This can be understood by analogy to $\matr{K}$. The off-diagonal and
diagonal elements of $\matr{K}$ correspond to transitions arriving at or departing from,
respectively, the promoter state of interest. If transcription initiation is 
accompanied by promoter state changes, we must have separate matrices for
arriving and departing transcription events since the arriving and departing
transitions have different initial copy numbers of mRNA, unlike for $\matr{K}$
where they are the same. The master equation for this model is
\begin{equation}
\deriv{t}\vect{p}_m(t) =
\left( \matr{K} - \matr{R_D} - \gamma m \matr{I} \right) \vect{p}_m(t)
                + \matr{R_A} \vect{p}_{m-1}(t) +
                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
\label{eq:3state_rep_cme}
\end{equation}
with the state vector and promoter transition matrix defined as
\begin{equation}
\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\ p_{m,P} \end{pmatrix},\
\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 \\
                        k_R^- & -k_R^+ -k_P^+ & k_P^- \\
                        0 & k_P^+ & -k_P^- 
                \end{pmatrix},
\label{eq:3state_cme_matrices_pt1}
\end{equation}
and the initiation matrices given by
\begin{equation}
\matr{R_A} = \begin{pmatrix}
                0 & 0 & 0 \\ 0 & 0 & r \\ 0 & 0 & 0
                \end{pmatrix},\
\matr{R_D} = \begin{pmatrix}
                0 & 0 & 0 \\ 0 & 0 & 0 \\ 0 & 0 & r
                \end{pmatrix},
\label{eq:3state_cme_matrices_pt2}
\end{equation}
The elements of $\vect{p}_m$ encode the probabilities of having $m$ mRNA present
along with the promoter having repressor bound, being empty, or having RNAP
bound, respectively. $\matr{R_A}$ describes probability flux \textit{arriving}
at the state $\vect{p}_m$ from a state with one fewer mRNA, namely
$\vect{p}_{m-1}$, and $\matr{R_D}$ describes probability flux \textit{departing}
from the state $\vect{p}_m$ for a state with one more mRNA, namely
$\vect{p}_{m+1}$. $\matr{K}$ is closely analogous to model 1.

Mean mRNA at steady state is found analogously to model 1, with the result
\begin{equation}
\langle m\rangle = \frac{r}{\gamma}
        \frac{k_R^- k_P^+}
        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)}.
\end{equation}
Fold-change is again found from the ratio prescribed by~\eq{eq:fc_def}, from
which we have
\begin{align}
\text{fold-change}
&=      \frac{k_R^- k_P^+}
        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)}
        \frac{k_P^+ + k_P^- + r}{k_P^+}
\\
&=      \left(1 + \frac{k_R^+}{k_R^-}
                \frac{k_P^- + r}{k_P^+ + k_P^- + r}
        \right)^{-1}
\\
&=      \left(1 + \frac{k_R^+}{k_R^-}
        \left(1 + \frac{k_P^+}{k_P^- + r}\right)^{-1}
        \right)^{-1},
\end{align}
which follows the master curve of~\fig{fig1:means_cartoons}(C) with $\rho = 1 + k_P^+/(k_P^- + r)$ as claimed.

\subsubsection{Nonequilibrium Model Three - Multistep Transcription Initiation and
Escape}
 One might reasonably complain that the first two ``nonequilibrium''
models we have considered are straw men. Their steady states necessarily satisfy
detailed balance which is equivalent to thermodynamic equilibrium. Why is this
the case? At steady state there is by definition no net probability flux in or
out of each promoter state, but since the promoter states form a linear chain,
there is only one way in or out of the repressor bound and RNAP bound states,
implying each edge must actually have a net zero probability flux, which is the
definition of detailed balance (usually phrased as equality of forward and
reverse transition fluxes).

Now we consider model 3 in Figure~\ref{fig1:means_cartoons}(B) which allows the possibility
of true nonequilibrium steady-state fluxes through the promoter states. We point
out that this model was considered previously in~\cite{Mitarai2015} where a
comparison was made with model 1 as used in~\cite{Jones2014}. The authors
of~\cite{Mitarai2015} argued that the additional complexity is essential to
properly account for the noise in the mRNA distribution. We will weigh in on
both models later when we consider observables beyond fold-change.
